### 2014-05-21 paper.draft.tex
paper draft

DHTs not only serve good for storing peer swarms on P2P applications. These overlay networks are used on a high variety of Internet services. For example, CDNS [1, 2], application-level multicast, instant messaging, domain name services, web caching or distributed file systems. #TODO references for each example.

@@

On the other hand, DHT technologies are mainly evolving thowards the Internet. However, emerging networks like Communit Networks [??] are very similar to the Internet but also have some differences.

Community Networks raise from the dramatic drop of wireless network equipement cost over the last decade which has enabled end-users to build their own network independently from traditional telecomunications providers.

Community Network infraestructure has some distinctive characteristics from the one used to power the Internet. Most of its traffic flows through short-distance, low-throughput, high-latency wireless links, build with cheap low-end hardware.

The topology characteristics of the network are also a bit different from the Internet. On village-like environments, typical CN deployment may involve a single Access Point on the top-roof of the villange which where all the clients contect to. On cities, usually the topology is a mesh-like network. Therefore, in both cases the network tends to forms clusters of nodes with a very few hops between each other but interconnected by long-distance _intervilage_ wireless links.

Locality-aware applications are much important in CN because the cost of leaving _your neighberhood cluster_ is not cheap since it may involve going through several kilometers of unreliable microwave links, with some packet loss, jitter, high latency and low throughput.

In this paper, we study the locality of DHT overlay on the context of Community Networks. In the next section background on DHT locality is presented

Background
==========
Network proximity in Distributed Hash Tables has been extensiviely studied [1, 2, 3]. Three basic approaches have been suggested for exploring proximity in DHT protocols:

@@

Geographic Layout techniques are very effective for building overlays with a high degree of physical locality. However, we found that in the context of community networks it may not be the best choice. Network splits are common due to both, hardware malfunction (low-end hardware used) and humman errors during management operations (difficulty of operating a descentralized network). Because of network splits a perfect locality may lead to entire portions of the DHT ID space inaccessible. In this sense Proximity routing and Proximity neighbour selection provides better reliability since the Node IDs have nothing to do with the geographic location.

Intuitively proximity routing may not work as good as Proximity neighbour selection. Because the routing table is small (compared to the total number of nodes), there is not a lot of candidates to choose from. Also priorizing pysical locality over logical locality may actually lead to an increase on the number of DHT hops when looking for a particular ID.

We think that Proximity Neighbour Selection offers the best characteristics for deployment on CN. ID space is uniformly distributed over all the network making it robust, and good enough locality can still be obtained by choosing good candidates four routing table entries.

Mainline DHT
============
We based our DHT evaluation on Mainline DHT. MLDHT is a Kademlia-based Distributed Hash Table used by BitTorrent clients to find peers via the BitTorrent protocol. MLDHT is the largest DHT network in the Internet, meassures from 2013 show from 10 million to 25 million users, with a daily churn of at least 10 million[1].

Mainline DHT protocol is very simple, limited to four control messages:

@@

Their simplicity and the fact that is a widely deployed protocol makes its evaluation eassier, since the compleixity is low and there are several MLDHT implementations to choose from. We looked at libtorrent and PYMDHT.

libtorrent[1] was our first consideration, is a well known open source implementation of BitTorrent protocol and it has support for Mainline DHT. However, we found out that its public API provides little information about the routing state, only node IDs, not IP addresses or RTT. This is not surprising since it is mainly designed focused on BitTorrent aplications, not for routing characteristics evaluation. Also libtorrent implements a highly tunne and complicated routing algorithm with a lot of corner cases and optimizations making its behaviour less predictable[2].

@@

PYMDHT [1] is a flexible Python implementation of the Mainline DHT protocol, specifically designed for MDHT evaluation. This fact has made us choose this library over libtorrent, even though PYMDHT current state is experimental, the codebase has room for improvement and there is no documentation available.

@@

Mainline DHT as described in Bittorrent Enhacement Proposal 5 (BEP5) [1] does not contemplate means for locality. However, implementers of this DHT are free to choose whatever routing policy they wish to. In the case of PYMDHT it supports a plugin system for this routing policy. The library ships with an implementation of a Proximity Neighbour Selection routing policy defined by NICE[2], based on RTT meassurements, which is very interesting for our locality evaluation.

@@

MLDHT Locality Evaluation on Community Lab
=================
The evaluation has been performed in Community Lab [1], a Community Networks Testbed by the CONFINE project[2], is a global facility for experimentation with network technologies and services for community networks.

Interestingly, Community Lab topology has also the type of node clustering commonly seen in community network villages. In the case of Community Lab those are the nodes at UPC-Lab where only one forwarding device is between every node[1].

@@

Because of the reduced number of available nodes in our testbed (in relation to the massive size of DHTs real deployments) we have found that we are not able to reproduce meaningful results of the locality characteristics of MDHT. Therefore, we have crafted our experiments intentionaly trying to ___ rather than real workload behaviours.

@@

* Communitylab suffers from interconnection problems between community networks, also Internet connectivity is not widely available; considerably reducing the number of useful nodes. Because, nodes don't always see each other and since Internet is not always available it increases the complexity of deploying our experiment in those disconnected nodes beyond what we consider feasible for our project's schedule.
* Reasons for low number of nodes: offline, no internet and isolated, lack of public ips

@@

>>>>

However, DHTs not only serve good for storing peer swarms on P2P applications. These overlay 
networks are used on a large variety of Internet services. For example, CDNS [1, 2], 
application-level multicast, instant messaging, domain name services, web caching 
or distributed file systems. #TODO references for each example.

@@

On the other hand, DHT technologies are mainly evolving thowards the Internet. 
Even though emerging networks like Communit Networks [??] are based on the same 
fundamental technology than Internet, they also have their own differences and peculiarities.

Community Networks raise from the dramatic drop of wireless network equipement 
cost over the last decade which has enabled end-users to build their own network 
independently from traditional telecomunications providers. This has a dramatic 
impact on the network properties. Most of its traffic flows through short-distance, 
low-throughput, high-latency wireless links, build with cheap low-end hardware.

The topology characteristics of the network are also a bit different from the Internet. 
On village-like environments, typical CN deployment may involve a single Access Point 
on the top-roof of the villange from which all the clients contect to. On cities, 
usually the topology is a mesh-like network. Therefore, in both cases the network 
tends to forms clusters of nodes with a very few hops between each other but interconnected 
by long-distance intervilage wireless links.

Locality-aware applications become much important in CN because the cost of leaving 
_your neighberhood cluster_ is particularly expensive since it may involve going 
through several kilometers of unreliable microwave links, with more packet loss, 
more jitter, higher latency and lower throughput.

In this paper, we study the locality of DHT overlay on the context of Community 
Networks. In the next section background on DHT locality is presented.

Background
==========
Network proximity (locality) in Distributed Hash Tables has been extensiviely studied [1, 2, 3]. 
Three basic approaches have been suggested for exploring proximity in DHT protocols:

@@

Geographic Layout techniques are very effective for building overlays with a 
high degree of physical locality. However, we found that in the context of 
community networks it may not be the best choice. Network splits are common due to both, 
hardware malfunction (low-end hardware used) and humman errors during management 
operations (difficulty of operating a highly descentralized network). Because of network 
splits a perfect locality may lead to entire portions of the DHT ID space inaccessible. 
In this sense, Proximity Routing and Proximity Neighbour Selection provides much better 
reliability since the Node IDs have nothing to do with the geographic location.

Intuitively Proximity Routing may not work as good as Proximity Neighbour Selection. 
Because the routing table is small (compared to the total number of nodes), there 
are not a lot of candidates to choose from. Also priorizing pysical locality over 
logical locality may actually lead to an increase on the number of DHT hops for 
lookup operations.

We believe that Proximity Neighbour Selection offers the best characteristics for 
deployment on CN. ID space is uniformly distributed over the whole network, making 
it robust, while good enough locality can still be achieved by choosing good candidates 
as routing table entries.

Mainline DHT
============
We based our DHT evaluation on Mainline DHT (MLDHT). MLDHT is a Kademlia-based 
Distributed Hash Table used by BitTorrent clients to find peers via the BitTorrent 
protocol. MLDHT is the largest DHT network in the Internet, meassures from 2013 
show from 10 million to 25 million users, with a daily churn of at least 10 million[1].

Mainline DHT protocol is actually pretty simple, limited to four control messages:

@@

Its simplicity and the fact that it is a widely deployed protocol makes its evaluation 
study somehow eassier, since the compleixity is low and there are several MLDHT 
implementations to choose from. We looked at libtorrent and PYMDHT.

libtorrent[1] was our first consideration, is a well known open source implementation 
of BitTorrent protocol and it has support for Mainline DHT. However, we quickly found out 
that its public API provides little information about the DHT routing state, only 
providing the node ID of its entrie, not IP addresses or RTT. This is not surprising since its designe is 
focused on BitTorrent aplications, not on DHT routing characteristics evaluation. 
Also libtorrent implements a highly tunne and complicated routing algorithm with a 
lot of corner cases and optimizations making its behaviour less predictable[2].

@@

PYMDHT [1] is a flexible Python implementation of the Mainline DHT protocol, 
specifically designed for MDHT evaluation. This fact has made us choose this 
library over libtorrent. Even though PYMDHT current state is experimental, 
the codebase has room for improvement and there is no public documentation available.

@@

Mainline DHT as described in Bittorrent Enhacement Proposal 5 (BEP5) [1] does not 
contemplate means for locality. However, implementers of this DHT are free to choose 
whatever routing policy they wish to. In the case of PYMDHT it supports a plugin system 
for the routing policy. The library ships with an implementation of a Proximity Neighbour 
Selection routing policy defined by NICE[2], based on RTT meassurements, which 
is very interesting for our locality evaluation.

@@

MLDHT Locality Evaluation on Community Lab
==========================================
The evaluation has been performed in Community Lab [1], a Community Networks Testbed 
by the CONFINE project[2], is a global facility for experimentation with network 
technologies and services for community networks.

Interestingly, Community Lab topology has also the type of node clustering commonly 
seen in community network villages. In the case of Community Lab those are the nodes 
at UPC-Lab where only one forwarding device sits between every node[1].

@@

Because of the reduced number of available nodes in our testbed (in relation to 
the massive DHT size in real deployments) we have found that we are not able to 
reproduce meaningful results of the locality characteristics of MLDHT. 
Therefore, we have crafted some workloads trying to artificaly promote locality 
rather than focus on real workload behaviours.

@@

* Communitylab suffers from interconnection problems between community networks, also Internet connectivity is not widely available; considerably reducing the number of useful nodes. Because, nodes don't always see each other and since Internet is not always available it increases the complexity of deploying our experiment in those disconnected nodes beyond what we consider feasible for our project's schedule.
* Reasons for low number of nodes: offline, no Internet and isolated, lack of CN public ips

@@

Naive changes on the routing table policy in order to improve locality may actually 
harm locality because of the nature of the MLDHT. More hops? disrupt logical locality?
